An Executable Formal Semantics For A Functional Actor Language

### 2017-12-20 paper/paper.tex
minor

%
\begin{document}
%
\mainmatter              % start of the contributions
%
\title{An Executable Formal Semantics For A Functional Actor Language}
%
\author{Xiaohong Chen, Lucas Pe\~{n}a}
%
\institute{University of Illinois at Urbana-Champaign, \\
\email{\{xc3, lpena7\}@illinios.edu}}

@@

\begin{figure}
          \centering
          \includegraphics[scale=0.13]{../k-overview.png}
	  \caption{Backend tools K provides from just an operational
            semantics. Some tools such as a semantics-based compiler and
            test-case generation are in progress.}
          \label{fig:k-tools}
\end{figure}
To demonstrate the power of K, languages as complex as C, Java, and JavaScript
have been fully defined using the K framework~\cite{kc,kjava,kjs}. On top of that, in
all of these languages and more, nontrivial programs have been formally verified
using K's prover, such as proofs involving AVL trees and red-black trees. We
would like to add an actor-based language to the growing list of languages
defined in K, so we can take advantage of K's formal verification engine and
more.

@@

\subsection{Semantics}\label{sec:semantics}
Semantics in K is given as \emph{rewriting rules} of the form $l \Rightarrow r$.
For example, the  beta-reduction rule (beta-v) is specified as
\begin{verbatim}
  rule app(lambda(X:Id, E:Exp), V:Val) => E [ V / X ]
\end{verbatim}
where \texttt{E [ V / X ]} is the result of substituting \texttt{V} for 
\texttt{X} in \texttt{E} using K's internal substitution mechanism.
Conditional reduction rules can also be easily specified.
For example, the branching rule (br) is specified as two K rules
\begin{verbatim}
  rule br(nil, V1:Val, V2:Val) => V2
  rule <k> br(V:Val, V1:Val, V2:Val) => V1 ... </k>
       <definedaddr> X </definedaddr>
    requires notBool((isId(V) andBool notBool(V in X))
                              orBool V ==K nil)
\end{verbatim}
The first rule says if the first argument is \texttt{nil}, the branching 
operator \texttt{br} returns its third argument.
The second rule is a bit complicated.
It says if the first argument is neither \texttt{nil} nor 
unknown addresses, the branching operator \texttt{br} returns its second 
argument.

The single-step transition relation $\mapsto$ on actor configurations is 
similarly specified using K rules.
For example, the \texttt{<init: a, a1>} transition rule is specified as
\begin{verbatim}
  rule <actor>
         <k> initbeh(A1, V:Val) => nil ... </k>
         <id> A </id>
       </actor>
       <actor>
         <k> uninit(A) => ready(V) ... </k>
         <id> A1 </id>
       </actor>
\end{verbatim}
The rule involves two actors as there are two actor cells 
\texttt{<actor>}\ \dots\ \texttt{</actor>} in the rule.
The first actor is the \emph{initializer} actor while the second is the actor 
\emph{being initialized}.
K supports to only specify components that are relevant to the rule. This is 
known as the \emph{framing principle}.
The initializer actor executes \texttt{initbeh(A1, V:Val)} and gets value 
\texttt{nil}, as specified in the rule.
The rest computation or \emph{the reduction context} is unchanged and kept in 
the following~``\texttt{...}''.

Reduction context can be mentioned explicitly when needed.
For example in the following \texttt{<bec: a, a1>} rule,
\begin{verbatim}
  rule <actor>
         <k> (become(V:Val) ~> RestComputation) => ready(V) </k>
         ...
       </actor>
       (.Bag => <actor>
                  <k> nil ~> RestComputation </k>
                  <id> !A1:Id </id>
                </actor>)
       <definedaddr> Ads => Ads ListItem(!A1) </definedaddr>
\end{verbatim}
the reduction context \texttt{RestComputation} is explicitly passed to a 
newly-generated actor.
The tilde arrow expression \verb|T ~> R| 
means the decomposition \verb|R[T]| with reduction context \texttt{R} and redex 
\texttt{T}.

\subsection{In-depth Example}
We test our actor semantics on many examples. One of the more complicated test
cases is an example of tree recursion~\cite{actor}.  The problem is to calculate
the product of all leaves of a binary tree.  The problem is solved by a
divide-and-conquer algorithm which calculates the results for two subtrees and
multiplies the results.  The algorithm is implemented using a
\emph{join continuation} as follows.
\small
\begin{verbatim}
let(bind(sink,
  rec(lambda(b, lambda(m, become(b))))),
let(bind(ident,
  lambda(m, m)),
let(bind(joincont,
  rec(lambda(b, lambda(c, lambda(nargs, lambda(firstnum, lambda(num,
    if(eq?(nargs, 0),
       become(app(app(app(b, c), 1), num)),
       seq(become(sink),
           send(c, mult(firstnum, num))))))))))),
let(bind(treeprod,
  rec(lambda(b, lambda(self, lambda(m,
    seq(become(app(b, self)),
        if(isnat(tree(m)),
           send(cust(m), tree(m)),
           letactor(bind(newcust, app(app(app(joincont, cust(m)), 0), nil)),
             seq(send(self, pr(left(tree(m)), newcust)),
                 send(self, pr(right(tree(m)), newcust))))))))))),
let(bind(tr, 
  pr(pr(1, pr(2, pr(3, 4))), pr(pr(5, 6), 7))),
letactor(
  bind(tpactor, app(treeprod, tpactor)),
  bind(custactor, ident),
  send(tpactor, pr(tr, custactor))))))))
\end{verbatim}
\normalsize
Recall that our semantics initializes the initial actor with the behavior of 
executing the above lambda expression.
Therefore after five \texttt{let}-expressions, the initial actor creates two 
actors. One is the \texttt{tpactor} that calculates the tree product.
The other is the \texttt{custactor} that receives the result.
The initial actor sends \texttt{tpactor} a message containing a binary tree 
whose leaves are $1, 2, \dots, 7$, and the address of the customer actor 
\texttt{custactor}.
When the actor system terminates, we get the following configuration
\begin{verbatim}
  <T>
    <actors>
      <actor>
        <k> ready ( ... ) </k>
        <id> _93 </id>
      </actor>
      <actor>
        <k> ready ( ... ) </k>
        <id> _179 </id>
      </actor>
      .... ....
      <actor>
        <k> ready ( ... ) </k>
        <id> _45 </id>
      </actor>
      <actor>
        <k> exec ( nil ) </k>
        <id> initactor </id>
      </actor>
      <actor>
        <k> exec ( 5040 ) </k>
        <id> _46 </id>
      </actor>
      ......
      ......
    </actors>
    <messages> .MessageCellBag </messages>
    <definedaddr> ... </definedaddr>
  </T>
\end{verbatim}
In the configuration there are many idle actors generated during the 
computation. The customer actor, which in this execution has identity 
\texttt{\_46}, is holding the final product of 5040.

\section{Conclusion and Future Work}

In this paper we showed how to define the actor semantics from~\cite{actor} using the
K framework. We described in detail the syntax and semantics of the language in
K, as well as the configuration for our actor language. We also showed how we
can use K's strictness attributes to emulate the reduction contexts described
in~\cite{actor}. Finally, we discussed a complicated actor program, the tree product
example, and its encoding in K.

>>>>

%
\begin{document}
%
\mainmatter              % start of the contributions
%
\title{An Executable Formal Semantics For A Functional Actor Language}
%
\author{Xiaohong Chen, Lucas Pe\~{n}a}
%
\institute{University of Illinois at Urbana-Champaign, \\
\email{\{xc3,lpena7\}@illinios.edu}}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{../k-overview.png}
  \caption{Backend tools K provides from just an operational
           semantics. Some tools such as a semantics-based compiler and
           test-case generation are in progress.}
  \label{fig:k-tools}
\end{figure}
To demonstrate the power of K, languages as complex as C, Java, and JavaScript
have been fully defined using the K framework~\cite{kc,kjava,kjs}. On top of that, in
all of these languages and more, nontrivial programs have been formally verified
using K's prover, such as proofs involving AVL trees and red-black trees. We
would like to add an actor-based language to the growing list of languages
defined in K, so we can take advantage of K's formal verification engine and
more.

@@

\subsection{Semantics}\label{sec:semantics}
Semantics in K is given as \emph{rewriting rules} of the form $l \Rightarrow r$.
For example, the  beta-reduction rule (beta-v) is specified as
\begin{verbatim}
  rule app(lambda(X:Id, E:Exp), V:Val) => E [ V / X ]
\end{verbatim}
where \texttt{E [ V / X ]} is the result of substituting \texttt{V} for 
\texttt{X} in \texttt{E} using K's internal substitution mechanism.
K rules start with the keyword \texttt{rule}. The left hand side of the rule 
and right hand side of the rule are separated by a double arrow~``\texttt{=>}''.

Conditional reduction rules can also be easily specified.
For example, the branching rule (br) is specified as two K rules
\begin{verbatim}
  rule br(nil, V1:Val, V2:Val) => V2
  rule <k> br(V:Val, V1:Val, V2:Val) => V1 ... </k>
       <definedaddr> X </definedaddr>
    requires notBool((isId(V) andBool notBool(V in X))
                              orBool V ==K nil)
\end{verbatim}
The first rule says if the first argument is \texttt{nil}, the branching 
operator \texttt{br} returns its third argument.
The second rule is a bit complicated.
It says if the first argument is neither \texttt{nil} nor 
unknown addresses, the branching operator \texttt{br} returns its second 
argument.
The \texttt{requires} keyword is used to specify rewriting conditions that 
decide when a rewrite rule can be applied.

The single-step transition relation $\mapsto$ on actor configurations is 
similarly specified using K rules.
For example, the \texttt{<init: a, a1>} transition rule is specified as
\small
\begin{verbatim}
  rule <actor>
         <k> initbeh(A1, V:Val) => nil ... </k>
         <id> A </id>
       </actor>
       <actor>
         <k> uninit(A) => ready(V) ... </k>
         <id> A1 </id>
       </actor>
\end{verbatim}
\normalsize
The rule involves two actors as there are two actor cells 
\texttt{<actor>}\ \dots\ \texttt{</actor>} in the rule.
The first actor is the \emph{initializer} actor while the second is the actor 
\emph{being initialized}.
K supports to only specify components that are relevant to the rule. This is 
known as the \emph{framing principle}.
The initializer actor executes \texttt{initbeh(A1, V:Val)} and gets value 
\texttt{nil}, as specified in the rule.
The rest computation or so-called \emph{reduction context} is unchanged and 
kept in 
the following~``\texttt{...}''.

Reduction context can be mentioned explicitly when needed.
For example, in the following \texttt{<bec: a, a1>} rule,
\small
\begin{verbatim}
  rule <actor>
         <k> (become(V:Val) ~> RestComputation) => ready(V) </k>
         ...
       </actor>
       (.Bag => <actor>
                  <k> nil ~> RestComputation </k>
                  <id> !A1:Id </id>
                </actor>)
       <definedaddr> Ads => Ads ListItem(!A1) </definedaddr>
\end{verbatim}
\normalsize
the reduction context \texttt{RestComputation} is explicitly passed to a 
newly-generated actor with a fresh identity \verb|!A1|.
The tilde arrow expression \verb|T ~> R| 
means the decomposition \verb|R[T]| with reduction context \texttt{R} and redex 
\texttt{T}.

\subsection{In-depth Example}
We test our actor semantics on many examples. One of the more complicated test
cases is an example of tree recursion~\cite{actor}.  The problem is to calculate
the product of all leaves of a binary tree.  The problem is solved by a
divide-and-conquer algorithm which calculates the results for two subtrees and
multiplies the results.  The algorithm is implemented using a
\emph{join continuation} as follows.
\small
\begin{verbatim}
let(bind(sink,
  rec(lambda(b, lambda(m, become(b))))),
let(bind(ident,
  lambda(m, m)),
let(bind(joincont,
  rec(lambda(b, lambda(c, lambda(nargs, lambda(firstnum, lambda(num,
    if(eq?(nargs, 0),
       become(app(app(app(b, c), 1), num)),
       seq(become(sink),
           send(c, mult(firstnum, num))))))))))),
let(bind(treeprod,
  rec(lambda(b, lambda(self, lambda(m,
    seq(become(app(b, self)),
        if(isnat(tree(m)),
           send(cust(m), tree(m)),
           letactor(bind(newcust, app(app(app(joincont, cust(m)), 0), nil)),
             seq(send(self, pr(left(tree(m)), newcust)),
                 send(self, pr(right(tree(m)), newcust))))))))))),
let(bind(tr, 
  pr(pr(1, pr(2, pr(3, 4))), pr(pr(5, 6), 7))),
letactor(
  bind(tpactor, app(treeprod, tpactor)),
  bind(custactor, ident),
  send(tpactor, pr(tr, custactor))))))))
\end{verbatim}
\normalsize

Recall that our semantics initializes the initial actor with the behavior of 
executing the above lambda expression.
Therefore after five nested \texttt{let}-expressions, the initial actor creates 
two 
actors. One is the \texttt{tpactor} that calculates the tree product.
The other is the \texttt{custactor} that receives the result.
The initial actor sends \texttt{tpactor} a message containing a binary tree 
whose leaves are $1, 2, \dots, 7$, and the address of the customer actor 
\texttt{custactor}.
When the actor system terminates, we get the following configuration
\small
\begin{verbatim}
  <T>
    <actors>
      <actor>
        <k> ready ( ... ) </k>
        <id> _93 </id>
      </actor>
      <actor>
        <k> ready ( ... ) </k>
        <id> _179 </id>
      </actor>
      .... ....
      <actor>
        <k> ready ( ... ) </k>
        <id> _45 </id>
      </actor>
      <actor>
        <k> exec ( nil ) </k>
        <id> initactor </id>
      </actor>
      <actor>
        <k> exec ( 5040 ) </k>
        <id> _46 </id>
      </actor>
      ......
      ......
    </actors>
    <messages> .MessageCellBag </messages>
    <definedaddr> ... </definedaddr>
  </T>
\end{verbatim}
\normalsize
In the configuration there are many idle actors generated during the 
computation. The customer actor, which in this execution has identity 
\texttt{\_46}, is holding the final product of 5040.

\section{Conclusion and Future Work}
\label{sec:conclusion}
In this paper we showed how to define the actor semantics from~\cite{actor} using the
K framework. We described in detail the syntax and semantics of the language in
K, as well as the configuration for our actor language. We also showed how we
can use K's strictness attributes to emulate the reduction contexts described
in~\cite{actor}. Finally, we discussed a complicated actor program, the tree product
example, and its encoding in K.
